Biodiversity temporal trends are reshaping food web structure and redundancy in riverine ecosystems

### 2025-04-10 paper/manuscript.Rmd
Update manuscript with Colin and Elisa comments

HERE

The total effect of the temporal trends in species richness on the temporal
trend in connectance was negative ($r_\delta =$ `r rttps$ct$r`, Fig.
\@ref(fig:std)a), resulting from a negative direct effect ($r_\delta =$
`r rdtps$ct$r`, Fig. \@ref(fig:sem)f) and a positive indirect effect via changes
in the temporal trends in community biomass ($r_\delta =$ `r ridtps$ct$r`).
Thus, when communities experienced an increase, or decrease, in species
richness, they also experienced a concomitant decrease, or increase, in
connectance. In contrast, species richness temporal trends had no direct effects
on trends in average trophic level ($r_\delta =$ `r rdtps$tlvl$r`, Fig.
\@ref(fig:sem)g), but a positive indirect effect via changes in the trends of
community biomass ($r_\delta =$ `r ridtps$tlvl$r`), which
adds up to a total positive effect of species richness temporal trends on
average trophic level ($r_\delta =$ `r rttps$tlvl$r`, Fig. \@ref(fig:std)a).
Temporal trends in species richness had a strong total positive effect on the
trends in redundancy (resp. $r_\delta =$ `r rdtps$red$r`, `r ridtps$red$r`), via
direct and indirect positive effects ($r_\delta =$ `r rttps$red$r`). All
together, this means that communities that have experienced species richness
declines over the last decades have decreased in average trophic level and
redundancy but increased in connectance. Conversely, those that experienced
species richness increases have increased in average trophic level, redundancy,
but decreased in connectance.

The effects of species richness were more complex, because species richness in
this model has indirect effects on food web structure through community biomass.
Species richness had a direct negative effect on connectance ($r_\delta =$
`r rdtps$ct$r`, Fig. \@ref(fig:sem)f), a positive indirect effect through
community biomass ($r_\delta =$ `r ridtps$ct$r`), so that species richness had a
negative effect in total on connectance ($r_\delta =$ `r rttps$ct$r`, Fig.
\@ref(fig:std)a). In contrast, species richness had no direct effects on
average trophic level ($r_\delta =$ `r rdtps$tlvl$r`, Fig. \@ref(fig:sem)g), a
positive indirect effect through community biomass ($r_\delta =$
`r ridtps$tlvl$r`), which adds up to a positive total effect of species richness
temporal trends on average trophic level ($r_\delta =$ `r rttps$tlvl$r`, Fig.
\@ref(fig:std)a). Species richness temporal trends had direct and indirect
positive effect on the temporal trends in redundancy (resp. $r_\delta =$
`r rdtps$red$r`, `r ridtps$red$r`), which added up to a strong positive effect
($r_\delta =$ `r rttps$red$r`).

Spatial analyses, either using empirical data or the bioenergetic food web
model, largely aligned with temporal trends (Fig. \@ref(fig:std)a and S4),
showing consistent positive effects of species richness on community biomass and
of community biomass on average trophic level and redundancy. Species richness
also exhibited a direct negative effect on connectance and a positive effect on
redundancy, mirroring temporal patterns. However, spatial and temporal analyses
diverged in their effects of species richness on average trophic level, with
spatial data showing a negative effect and temporal trends indicating a positive
one. Detailed statistical effects and comparisons are provided in the Appendix.

@@

We implemented three secondary extinction simulation scenarios on the empirical
metaweb defined as a food web containing all the trophic species, i.e. the fish
species subdivided in size classes added to their resources, found across all
sites in the empirical data. This was to access a second line of inference about
whether top species were drivers of biomass and structure patterns (see section
4.2). When species are removed in a sequential order beginning by the highest
trophic levels, connectance, average trophic level, and redundancy decreased
(purple lines, Fig. \@ref(fig:std)b). On the other hand, removing low trophic level species
first increases connectance and average trophic level, but did not affect
redundancy (yellow lines, Fig. \@ref(fig:std)b). In turn, random species extinction sequence
had little effect on food web structure (black line, Fig. \@ref(fig:std)b). These results,
in conjunction with our empirical analyses above, suggest that top consumers
play a disproportionate role in driving the changes in food web structure
reported.

```{r}
std_cap <- paste0( "a, Direct and total effects of community biomass and species
  richness trends on food web structure trends. Standardized coefficients (r)
  derived from the temporal, spatial, and bioenergetic model Structural Equation
  Models. Points and lines display the estimated effects and their confidence
  intervals at 95\\%. Values are displayed in Table S4, R-squared in Table S5, and
  the visualization of the bivariate relationships for the spatial and
  bioenergetic model SEMs are displayed in Fig. S4. Average trophic level is
  weighted by the biomass of the trophic species. Redundancy is the proportion
  of redundant links. b, Effect of species extinctions on the structure of the
  empirical metaweb according to three scenarios. In the Decreasing and
  Increasing scenarios, species were ordered by their trophic level and removed
  from the metaweb starting respectively from the highest and lowest trophic
  level. Species were removed in a random sequence (50 sequences, grey dots) for
  the random scenario, the black line displaying the average predicted by a GAM
  linear model. For the extinction sequence, the average trophic level was not
  weighted by the biomass of species as this information is not available at the
  metaweb scale. Similar results were obtained with trophic species (i.e. node)
  extinction (Fig. S5).")
```

```{r std, fig.cap=std_cap, out.width = "100%"}
tar_load(p_sem_tot_ext_sp_file)
knitr::include_graphics(p_sem_tot_ext_sp_file[[1]])
```

@@

Our findings reveal several key patterns: (1) temporal trends in species
richness were positively associated with trends in community biomass; (2)
temporal trends in community biomass were positively linked to trends in
connectance and average trophic level; (3) temporal trends in species richness
were negatively associated with trends in connectance but only weakly related to
trends in average trophic level; and (4) both community biomass and species
richness trends were positively associated with trends in trophic pathway
redundancy. Additionally, we found that (5) top predator species play a
significant role in driving these patterns. Importantly, these temporal
relationships were largely consistent with patterns observed across spatial
gradients, both in empirical data and in simulations from a bioenergetic food
web model. In the following discussion, we explore these findings in the context
of food web assembly theory and Biodiversity-Ecosystem Functioning theory, and
consider their broader ecological implications.

We found that community biomass scaled positively with species richness, a
pattern consistent across our temporal, spatial, and theoretical analyses. This
result aligns with previous studies in fish communities [@woods_testing_2020;
@danet_species_2021] as well as in grasslands, salt marshes, and forests
[@hector_plant_1999; @tilman_biodiversity_2014; @wu_relationship_2015;
@li_relationship_2018; but see @grace_integrative_2016]. These empirical
findings are supported by theory, which suggests that food web diversity
enhances biomass by stimulating primary production and resource uptake
[@schneider_animal_2016; @wang_biodiversity_2018; @wang_intraguild_2019;
@nie_will_2023] and by Biodiversity-Ecosystem Functioning theory, which predicts
that species loss reduces community biomass [@loreau_biodiversity_2001]. Our
results reinforce the idea that declines in species richness can lead to reduced
biomass across ecosystems, highlighting the potential for conservation actions
that promote species richness to also enhance community biomass.

Our results also reveal that food webs experiencing biomass declines undergo
significant structural changes, including the loss of top predators and reduced
connectance. These findings align with energetic and trophic biogeography
theories [@lindeman_trophic-dynamic_1942; @gravel_trophic_2011], which posit
that the presence of high-trophic-level species is constrained by the
availability of energy and resources [@lindeman_trophic-dynamic_1942,
@takimoto_ecosystem_2008; @post_ecosystem_2000; @post_long_2002]. Thus, a
decline in community biomass may lead to insufficient resources to sustain top
predators, while an increase in biomass could provide the necessary energy to
support these species [@takimoto_ecosystem_2008; @post_ecosystem_2000].
Alternatively, Biodiversity-Ecosystem Functioning theory suggests that the
presence of higher trophic levels may reduce niche overlap among producers,
enhancing productivity and community biomass [@schneider_animal_2016;
@wang_biodiversity_2018; @wang_intraguild_2019; @nie_will_2023]. The
relationship between community biomass and connectance is less clear, with a
weaker positive association compared to average trophic level. Previous
theoretical studies have reported contrasting effects of connectance on
productivity [@nie_will_2023; @wang_intraguild_2019; @thebault_food-web_2003;
@poisot_trophic_2013; @schneider_animal_2016]. For instance, while increased
connectance between herbivores and plants can either enhance or reduce
productivity [@thebault_food-web_2003; @poisot_trophic_2013; @nie_will_2023],
greater connectance among carnivores generally boosts productivity by reducing
top-down control and allowing biomass to accumulate across trophic levels
[@wang_intraguild_2019; @schneider_animal_2016].

Our study demonstrates that temporal trends in community biomass and species
richness are positively associated with the proportion of redundant trophic
pathways in food webs. In other words, declines in community biomass and species
richness lead to a loss of redundant links, while increases in these factors
result in greater redundancy. Since redundant links are those that can be
removed without triggering secondary extinctions [@allesina_who_2004;
@bodini_using_2009], our findings suggest that declines in community biomass and
species richness reduce food web robustness to further perturbations
[@allesina_functional_2009]. The positive effect of community biomass on
redundancy aligns with theoretical predictions: higher redundancy can reduce
trophic complementarity within species, weaken top-down control, and promote
compensatory dynamics, thereby enhancing robustness following extinctions
[@staniczenko_structural_2010]. Similarly, the positive effect of species
richness on redundancy supports the insurance hypothesis
[@yachi_biodiversity_1999; @loreau_biodiversity_2021], where greater species
richness and functional redundancy buffer ecosystems against perturbations,
increasing stability [@sanders_trophic_2018].

Furthermore, our results highlight the disproportionate role of top
trophic-level species in maintaining redundant links within food webs. This
pattern is consistent with previous empirical studies showing that top predators
are often the most generalist and highly connected species
[@montoya_ecological_2006; @montoya_press_2009]. Consequently, disturbances
affecting these top predators are likely to have the greatest impact on overall
community stability [@montoya_press_2009]. These findings underscore the
importance of preserving top predators and maintaining species richness to
safeguard food web robustness and ecosystem stability.

Species richness was negatively associated with connectance, consistent with
previous findings and likely driven by limitations in predator generality
[@dunne_network_2006]. In contrast, the effects of species richness on average
trophic level were more nuanced and considerably weaker than those of community
biomass. Direct effects of species richness on average trophic level were either
negligible or negative, as observed in temporal, spatial, and theoretical
models. However, when accounting for the indirect effects of species
richness—mediated through changes in community biomass—temporal and theoretical
analyses revealed a positive association between species richness and average
trophic level. These contrasting effects may be explained by our extinction
simulations, which showed that the removal of top trophic-level species had a
much smaller impact on average trophic level than on connectance. This suggests
that while species richness influences food web structure, its effects are more
pronounced on connectance than on average trophic level dynamics.

We found strong agreement between our temporal and spatial analyses, as well as
with predictions from food web theory and Biodiversity-Ecosystem Functioning
research. This suggests that spatial comparisons of generic topological food web
properties and biodiversity metrics may serve as reasonable surrogates for
predicting temporal trends—supporting the validity of the space-for-time
substitution in both empirical data and bioenergetic food web models. However,
this agreement did not hold for the relationship between weighted average
trophic level (hereafter “average trophic level”) and species richness. This
discrepancy may arise because average trophic level incorporates both food web
topology and biomass distribution, which are influenced by dynamic processes
such as bottom-up and top-down control. These processes likely require
additional information about species metabolism and productivity
[@brose_allometric_2006; @barbier_pyramids_2019]. This interpretation is
supported by the fact that our spatial analysis accurately predicted the
temporal association between community biomass and average trophic level.
Importantly, this finding highlights that the relationship between
biodiversity—biomass and species richness—and food web
characteristics—connectance, trophic level, and redundancy—is not purely
structural but also functional. Indeed, predictions of food web structure are
most accurate when both structural (species richness) and functional (biomass)
metrics are considered.

Our study provides compelling evidence that biodiversity trends are profoundly
reshaping food web structure. By linking these patterns to
Biodiversity-Ecosystem Functioning and food web theory, we demonstrate that
widespread biodiversity declines likely have significant and detrimental
consequences for ecosystem functioning and stability in the face of future
perturbations. These findings suggest that assessments of food web structure,
community biomass, and species richness—which are often more feasible than
direct measurements of ecosystem functions—could serve as valuable indicators of
functional decline and metrics for evaluating the efficacy of mitigation
strategies. Importantly, our study also highlights critical avenues for future
research, stemming from its limitations. For instance, integrating food web
metrics into conservation planning will require understanding how they relate to
other facets of biodiversity, such as the presence of non-native species, and
drivers of biodiversity changes, including habitat degradation
[@danet_past_2024] and climate change [@bonnaffe_interaction_2024]. Ultimately, our
work aims to bridge the gap between Biodiversity-Ecosystem Functioning and food
web theory, fostering greater integration with research on biodiversity changes.

@@

We first estimated the body mass of each fish individual with the allometric
law: $B_i = 0.01 \times l_i^{3.03}$, $B_i$ being the biomass expressed in gram and $l_i$ the
body length of the individual $i$ expressed in millimeters [@froese_cube_2006;
@danet_species_2021]. Then the biomass of each species and community
were obtained by summing the body mass of individuals. This resulted in a
biomass score for the whole community of each site. Finally, the biomass was
reported to the sampled surface (in $m^2$) to account for sampling effort.

@@

We inferred the metaweb according to ontogenic diet data and body size
[@gravel_inferring_2013; @poisot_trophic_2013; @brose_predator_2019;
@bonnaffe_comparison_2021; @danet_species_2021]. The method is extensively
described in @bonnaffe_comparison_2021 and @danet_species_2021. The food diet of
stream fishes changes along their life cycle or stage, therefore according to
their body size. Each fish species was divided in nine size classes, i.e.
trophic species, equally distributed in their observed body size range. Thus,
a trophic species is a group of individuals of the same species and size class
that are expected to share the same prey and predators because trophic
interactions are largely determined by predator-prey size ratio
[@brose_allometric_2006; @brose_predator_2019]. We also added seven nodes present in the ontogenic
food diet database: detritus, biofilm, phytoplankton, zooplankton, macrophages,
phytobenthos and zoobenthos.  The resulting metaweb contained 45 fish species
divided into 9 size classes plus 7 resource nodes, summing up to 412 nodes.
Consideration of size classes, i.e.  trophic species instead of species,
improves inference of trophic interactions [@jennings_weak_2001].

@@

At each site, yearly food webs were described by their community biomass and
species richness. Yearly food web structure was described by their connectance,
describing the degree of generalism in a network [@dunne_network_2006;
@delmas_analysing_2019]. Connectance was computed as the ratio between the
number of actual links and the number of potential links ($C =
\frac{L}{S(S-1)}$, with $L$ being the number of links and $S$ the number of
trophic species). We also computed the average trophic level, weighted by the
biomass of the trophic species. The trophic level of a trophic species was
computed recursively as 1 added to the mean trophic level of the preys it
consumes [@herendeen_ecological_2008]. We computed the number of redundant and
functional links using the dominator tree approach [@allesina_functional_2009].
A species $i$ is a dominator of species $j$ if $i$ mediates the connection of $j$
to basal resources such that the extinction of $i$ removes the connection of $j$ to
basal resources. It is then used to identify the set of trophic links that are
essential to connect species to basal resources, i.e. functional links, and the
redundant links. This method requires a single basal root in the food webs. To
do so, we added a basal dummy resource node to the empirical food webs connected
to biofilm and phytoplankton resource nodes. For food webs generated from
theoretical simulations (see below), we added a resource node connected to all
primary producers. We computed the set of dominator trees using the
`dominator_tree()` function from the `igraph` R package (Appendix B). We
computed the number of functional links as the number of links in the resulting
dominator tree of the food web. We checked that the procedure reproduced the
examples presented in previous studies [@allesina_functional_2009, Appendix B].
We then derived the proportion of redundant links in food webs of empirical and
theoretical simulations.

@@

We modeled the temporal trends of food web structure at each site ($Y_i$) using
linear hierarchical model which accounts for the hierarchical spatial structure
of the data, i.e. the sites nested in hydrographic basins. Time was considered as the number of years since the
first sampling ($t = 0$) at each site. $\alpha$ is the intercept and $\beta$ the
slope estimating the temporal trends. The intercept and the slope parameters
have a fixed ($\alpha_0$, $\mu$) part and a random part. We accounted for the
spatial structure of the data by adding random intercepts ($a$) and slopes ($b$)
for the basin ($a_n$, $b_n$) and site ($a_{i|n}$, $b_{i|n}$) identity, site
effects being nested in the basin one.

@@

We extracted the temporal trends at each site from the hierarchical model
using mean BLUP values (Best Linear Unbiased Prediction), then multiplied
by 10 to get temporal trends by decade. For community biomass and species
richness which were log-transformed, we derived temporal trends in percentage
change by decade by back-transforming the slopes ($\beta\prime = (e^{\beta } - 1)
\times 100 \times 10$).

Overall, we found that standard deviation attributed to site effects in both
baseline and temporal trends in community structure were much higher than the
ones found for hydrographic basin (Table S6).

@@

The coefficients of the Structural Equation Models were computed with
PiecewiseSEM R package [@lefcheck_piecewisesem_2016]. The coefficients were
standardized and the confidence intervals were computed with semEff R package
[@murphy_semeff_2021]. Confidence intervals were obtained by bootstrapping the
linear models 1000 times and computing adjusted bootstrap percentile intervals
(BCa), which are adjusted for the asymmetric distribution of resampling. The marginal
and conditional $R^2$ associated with the linear models of the SEMs were computed
with the Nagelkerke method [@lefcheck_piecewisesem_2016, Table S5].

All the statistical models displayed low multicollinearity (Variance inflation
factors < 3, Table S7). We visually inspected the residuals of the linear
models.

@@

We simulated extinction sequences in the metaweb according to three scenarios to
get expectations on how the food web structure should change according to the
addition and deletion of species and nodes. The scenario of extinction sequences
were related to the trophic level of the species/nodes. In the decreasing and
increasing scenario, we removed species/nodes in decreasing, resp. increasing
order, of their trophic level, i.e. respectively beginning by the highest and
lowest trophic level. In the random extinction sequence, we built 50 random
extinction sequences. For each simulation, we removed fish species/nodes that
lost all their prey and iterate this removal until all remaining species have at
least one prey. We then computed the connectance, the average and the maximum
trophic level after each extinction.

@@

Consumers gain biomass by eating other species and lose biomass by being
consumed and by metabolic losses ($x_i$, eq. \@ref(eq:cons)). The functional
response (eq. \@ref(eq:holling)) of consumer feeding depends on the preference
for its preys ($\omega_{ij}$, set even across consumer preys) and
half-saturation rate ($B_0$, set to 0.5). The hill coefficient ($h$, set to 1.5)
determines the shape of the functional response (Holling's type II or III).

>>>>

The total effect of the temporal trends in species richness on the temporal
trend in connectance was negative ($r_\delta =$ `r rttps$ct$r`, Fig.
\@ref(fig:std)a), resulting from a negative direct effect ($r_\delta =$
`r rdtps$ct$r`, Fig. \@ref(fig:sem)f) and a positive indirect effect via changes
in the temporal trends in community biomass ($r_\delta =$ `r ridtps$ct$r`). Thus, when
communities experienced an increase, or decrease, in
species richness, they also experienced a concomitant decrease, or increase, in
connectance. In contrast, the total effects of the temporal trends in species
richness on the temporal trend of both trophic level and redundancy were
positive (resp. $r_\delta =$ `r rttps$tlvl$r`, `r rttps$red$r`, respectively,
Fig. \@ref(fig:std)a). The total positive effect on the temporal trend of
trophic level resulted from a combination of negative direct and positive
indirect effect while the effect on the temporal trend of redundancy resulted
from both positive direct and indirect effects (Fig. \@ref(fig:std)a). All together, this
means that communities that have experienced decline in species richness over
the last decades have decreased in average trophic level and redundancy but
increased in connectance. Conversely, those that experienced species richness
increases have increased in average trophic level, redundancy, but decreased in
connectance.

## Spatial association between biodiversity and food web

Spatial analyses, either using empirical data or the bioenergetic food web
model, largely aligned with temporal trends (Fig. \@ref(fig:std)a and S4),
showing consistent positive effects of species richness on community biomass and
of community biomass on average trophic level and redundancy. Across space,
sites with higher fish biomass on average had food webs with higher connectance,
redundancy and average trophic level than sites with lower fish biomass. The
theoretical model predicted similar relations, with a particularly strong
positive association between total biomass and average trophic level across
modelled food webs. Species richness also exhibited a direct negative effect on
connectance and a positive effect on redundancy, mirroring temporal patterns.
Both modelled food webs and local fish communities characterized by greater
species richness on average showed higher average biomass and food web
redundancy, and lower connectance. However, spatial and temporal analyses
diverged in their effects of species richness on average trophic level, with
spatial data showing a negative effect and temporal trends indicating a positive
one. Detailed statistical effects and comparisons are provided in the Appendix.

@@

To investigate whether top species were drivers of biomass and structure
patterns (see section 4.2), we implemented three secondary extinction simulation
scenarios on the empirical metaweb defined as a food web containing all the
trophic species, i.e. the fish species subdivided in size classes added to their
resources, found across all sites in the empirical data (See Methods). When
trophic species were removed in a sequential order beginning by the highest
trophic levels, connectance, average trophic level, and redundancy decreased
(purple lines, Fig. \@ref(fig:std)b). On the other hand, removing low level
trophic species first increases connectance and average trophic level, but does
not affect redundancy (yellow lines, Fig. \@ref(fig:std)b). In turn, species
extinction at random had little effect on food web structure (black line, Fig.
\@ref(fig:std)b). These results, in conjunction with our empirical analyses
above, suggest that top consumers play a disproportionate role in driving the
changes in food web structure reported.

(ref:fig-std-cap) a, Direct and total effects of community biomass and species
richness trends on food web structure trends. Standardized coefficients (r)
derived from the temporal, spatial, and bioenergetic model Structural Equation
Models. Points and lines display the estimated effects and their confidence
intervals at 95%. Values are displayed in Table S4, R-squared in Table S4, and
the visualization of the bivariate relationships for the spatial and
bioenergetic model SEMs are displayed in Figure S4. Average trophic level is
weighted by the biomass of the trophic species. Redundancy is the proportion of
redundant links. b, Effect of species extinctions on the structure of the
empirical metaweb according to three scenarios. In the Decreasing and Increasing
scenarios, species were ordered by their trophic level and removed from the
metaweb starting respectively from the highest and lowest trophic level. Species
were removed in a random sequence (50 sequences, grey dots) for the random
scenario, the black line displaying the average predicted by a GAM linear model.
For the extinction sequence, the average trophic level was not weighted by the
biomass of species as this information is not available at the metaweb scale.
Similar results were obtained with trophic species (i.e. node) extinction (Fig.
S5).

```{r std, fig.cap="(ref:fig-std-cap)", out.width = "100%"}
tar_load(p_sem_tot_ext_sp_file)
knitr::include_graphics(p_sem_tot_ext_sp_file[[1]])
```

@@

Our findings reveal several key patterns: (1) temporal trends in species
richness were positively associated with trends in community biomass; (2)
temporal trends in community biomass were positively linked to trends in
connectance and average trophic level; (3) temporal trends in species richness
were negatively associated with trends in connectance and only weakly related to
trends in average trophic level; and (4) both community biomass and species
richness trends were positively associated with trends in trophic pathway
redundancy. Additionally, we found evidence that (5) top predator species play a
significant role in driving these patterns. Importantly, these temporal
relationships were largely consistent with patterns observed across spatial
gradients, both in empirical data and in simulations from a bioenergetic food
web model. In the following, we discuss these findings in the context of food
web assembly theory and Biodiversity-Ecosystem Functioning theory, and consider
their broader ecological implications.

We found that community biomass scaled positively with species richness, a
pattern consistent across our temporal, spatial, and theoretical analyses. This
result aligns with previous studies on fish communities across space
[@woods_testing_2020; @danet_species_2021] as well as grasslands, salt marshes,
and forests [@hector_plant_1999; @tilman_biodiversity_2014;
@wu_relationship_2015; @li_relationship_2018; but see @grace_integrative_2016].
These empirical findings are supported by theory, suggesting that food web
diversity enhances biomass by stimulating primary production and resource uptake
[@schneider_animal_2016; @wang_biodiversity_2018; @wang_intraguild_2019;
@nie_will_2023], and by Biodiversity-Ecosystem Functioning theory, which
predicts that species loss reduces community biomass
[@loreau_biodiversity_2001]. Our results bring a temporal dimension to these
findings using long-term trends in empirical data, and clearly exemplify that
declines in species richness goes along with a reduction of biomass in
ecosystems. Conversely, our result also indicates that conservation actions
promoting species richness should also promote community biomass.

Our results also reveal that food webs experiencing biomass and/or species
richness declines undergo significant structural changes, including the loss of
top predators and a reduction of connectance. These findings align with
energetic and trophic biogeography theories [@lindeman_trophic-dynamic_1942;
@gravel_trophic_2011], which posit that the presence of high-trophic-level
species is constrained by the availability of energy and resources
[@post_ecosystem_2000]. Thus, a decline in community biomass may lead to
insufficient resources to sustain top predators, while an increase in biomass
could provide the necessary energy to support these species
[@takimoto_ecosystem_2008; @post_ecosystem_2000]. Further,
Biodiversity-Ecosystem Functioning theory suggests that the presence of higher
trophic levels may reduce niche overlap among producers, enhancing productivity
and community biomass [@schneider_animal_2016; @wang_biodiversity_2018;
@wang_intraguild_2019; @nie_will_2023]. The relationship between community
biomass and connectance is less clear, with a weaker positive association
compared to average trophic level. Previous theoretical studies have reported
contrasting effects of connectance on productivity [@nie_will_2023;
@wang_intraguild_2019; @thebault_food-web_2003; @poisot_trophic_2013;
@schneider_animal_2016]. For instance, while increased connectance between
herbivores and plants can either enhance or reduce productivity
[@thebault_food-web_2003; @poisot_trophic_2013; @nie_will_2023], greater
connectance among carnivores generally boosts productivity by reducing top-down
control and allowing biomass to accumulate across trophic levels
[@wang_intraguild_2019; @schneider_animal_2016]. Interestingly, our theoretical
model predicts overall a positive relation between total biomass and
connectance, which contrasts with the negative relation found recently by
another complex allometric food web model [@nie_will_2023], but this relation is
far weaker than the strong positive relation between total biomass and average
trophic level.

Our study suggests that declines in community biomass and species richness
are associated with a loss of redundant trophic links, while their increase
results in greater redundancy. Since redundant links are those that can be
removed without triggering secondary extinctions [@allesina_who_2004;
@bodini_using_2009], our findings suggest that declines in community biomass and
species richness might reduce food web robustness to further perturbations
[@allesina_functional_2009]. Such positive effect of species richness on
redundancy would then further support the insurance hypothesis
[@yachi_biodiversity_1999; @loreau_biodiversity_2021], where greater species
richness and functional redundancy buffer ecosystems against perturbations,
increasing stability [@sanders_trophic_2018]. Furthermore, our results highlight
the disproportionate role of top trophic-level species in maintaining redundant
links within food webs. This pattern is consistent with previous empirical
studies showing that top predators are often the most generalist and highly
connected species [@montoya_press_2009; @montoya_ecological_2006]. Consequently,
disturbances affecting these top predators are likely to have the greatest
impact on overall community stability [@montoya_press_2009]. These findings
underscore the importance of preserving top predators and maintaining species
richness to safeguard food web robustness and ecosystem stability.

Species richness was negatively associated with connectance, which is consistent
with previous findings and likely driven by limitations in predator generality
[@dunne_network_2006]. In contrast, the effects of species richness on average
trophic level were more nuanced and considerably weaker than those of community
biomass. Direct effects of species richness on average trophic level were either
negligible or negative, as observed in temporal, spatial, and theoretical
models. However, when accounting for the indirect effects of species richness —
mediated through changes in community biomass — our temporal and theoretical
analyses revealed a positive association between species richness and average
trophic level. This suggests that while species richness influences food web
structure, its effects are more pronounced on connectance than on average
trophic level dynamics.

We found strong agreement between our temporal and spatial analyses, as well as
with predictions from food web theory and Biodiversity-Ecosystem Functioning
research. This suggests that spatial comparisons of generic topological food web
properties and biodiversity metrics may serve as reasonable surrogates for
predicting temporal trends—supporting the validity of the space-for-time
substitution in both empirical data and bioenergetic food web models. However,
this agreement did not hold for the relationship between weighted average
trophic level (hereafter “average trophic level”) and species richness. This
discrepancy may arise because average trophic level incorporates both food web
topology and biomass distribution, which are influenced by dynamic processes
such as bottom-up and top-down control. Indeed, biomass distribution across
trophic levels depends on  species metabolism and productivity which in turn
control the strength of top-down and bottom-up processes
[@brose_allometric_2006; @barbier_pyramids_2019].

Our study provides compelling evidence that biodiversity trends are profoundly
reshaping food web structure. By linking these patterns to
Biodiversity-Ecosystem Functioning and food web theory, we demonstrate that
widespread biodiversity declines likely have significant and detrimental
consequences for ecosystem functioning and stability in the face of future
perturbations. These findings suggest that assessments of food web
structure—based on metaweb reconstruction like in the present study—, added to
more classic metrics of community biomass, and species richness could serve as
valuable indicators of functional decline and metrics for evaluating the
efficacy of mitigation strategies. Finally, our study also highlights critical
avenues for future research. For instance, integrating food web metrics into
conservation planning will require understanding how they relate to other facets
of biodiversity, such as the presence of non-native species, and drivers of
biodiversity changes, including habitat degradation [@danet_past_2024] and
climate change [@bonnaffe_interaction_2024].

@@

We first estimated the body mass of each fish individual with the allometric
law: $B_i = 0.01 \times l_i^{3.03}$, $B_i$ being the biomass expressed in gram
and li the body length of the individual $i$ expressed in millimeters
[@froese_cube_2006; @danet_species_2021]. Then the biomass of each species and
community were obtained by summing the body mass of individuals. This resulted
in a biomass score for the whole community of each site. Finally, the biomass
was reported to the sampled surface (in $m^2$) to account for sampling effort.

@@

We built the metaweb according to ontogenic diet data and body size
[@gravel_inferring_2013; @poisot_trophic_2013; @brose_predator_2019;
@bonnaffe_comparison_2021; @danet_species_2021; @bonnaffe_interaction_2024]. The
method is extensively described in @bonnaffe_comparison_2021 and
@danet_species_2021. The food diet of stream fishes changes along their life
cycle or stage, therefore according to their body size. Each fish species was
divided in nine size classes, i.e. trophic species, equally distributed in their
observed body size range. Thus, a trophic species is a group of individuals of
the same species and size class that are expected to share the same prey and
predators because trophic interactions are largely determined by predator-prey
size ratio [@brose_allometric_2006; @brose_predator_2019]. We also added seven
nodes present in the ontogenic food diet database: detritus, biofilm,
phytoplankton, zooplankton, macrophages, phytobenthos and zoobenthos. The
resulting metaweb contained 45 fish species divided into 9 size classes plus 7
resource nodes, summing up to 412 nodes. Consideration of size classes, i.e.
trophic species instead of species, improves inference of trophic interactions
[@jennings_weak_2001].

@@

At each site, yearly food webs were described by their community biomass and
species richness. Yearly food web structure was described by their connectance,
describing the degree of generalism in a network [@dunne_network_2006;
@delmas_analysing_2019]. Connectance was computed as the ratio between the
number of actual links and the number of potential links $C = \frac{L}{S(S-1)}$,
with $L$ being the number of links and $S$ the number of trophic species). We
also computed the average trophic level, weighted by the biomass of the trophic
species. The trophic level of a trophic species was computed recursively as 1
added to the mean trophic level of the preys it consumes
[@herendeen_ecological_2008]. We computed the number of redundant and functional
links using the dominator tree approach [@allesina_functional_2009]. A species
$i$ is a dominator of species $j$ if species $i$ mediates the connection of
species $j$ to basal resources such that the extinction of i removes the
connection of $j$ to basal resources. It is then used to identify the set of
trophic links that are essential to connect species to basal resources, i.e.
functional links, and the redundant links. This method requires a single basal
root in the food webs. To do so, we added a basal dummy resource node to the
empirical food webs connected to biofilm and phytoplankton resource nodes. For
food webs generated from theoretical simulations (see below), we added a
resource node connected to all primary producers. We computed dominator trees
using the `dominator_tree()` function from the `igraph` R package (See Appendix
for details). We computed the number of functional links as the number of links
in the resulting dominator tree of the food web. We checked that the procedure
reproduced the examples presented in previous studies
[@allesina_functional_2009, Appendix]. We then derived the proportion of
redundant links in food webs of empirical and theoretical simulations.

@@

We modeled the temporal trends of food web structure at each site ($Y_i$) using
a linear hierarchical model which accounts for the hierarchical spatial structure
of the data, i.e. the sites nested in hydrographic basins. Time was considered as the number of years since the
first sampling ($t = 0$) at each site. $\alpha$ is the intercept and $\beta$ the
slope estimating the temporal trends. The intercept and the slope parameters
have a fixed ($\alpha_0$, $\mu$) part and a random part. We accounted for the
spatial structure of the data by adding random intercepts ($a$) and slopes ($b$)
for the basin ($a_n$, $b_n$) and site ($a_{i|n}$, $b_{i|n}$) identity, site
effects being nested in the basin one.

@@

We extracted the annual temporal trends at each site from the hierarchical model
using mean Best Linear Unbiased Prediction (BLUP) values, then multiplied
by 10 to get temporal trends by decade. For community biomass and species
richness which were log-transformed, we derived temporal trends in percentage
change by decade by back-transforming the slopes ($\beta\prime = (e^{\beta } - 1)
\times 100 \times 10$).

Overall, we found that standard deviation attributed to site effects in both
baseline (i.e. the intercept) and temporal trends in community structure were much higher than the
ones found for hydrographic basin (Table S6).

@@

The coefficients of the Structural Equation Models were computed with
PiecewiseSEM R package [@lefcheck_piecewisesem_2016]. The coefficients were standardized and the
confidence intervals were computed with semEff R package [@murphy_semeff_2021]. Confidence
intervals were obtained by bootstrapping the linear models 1000 times and
computing adjusted bootstrap percentile intervals (BCa), which are adjusted for
the asymmetric distribution of resampling. The marginal and conditional R2
associated with the linear models of the SEMs were computed with the Nagelkerke
method [@lefcheck_piecewisesem_2016, Table S5].

All the statistical models displayed low multicollinearity (Variance Inflation
Factors < 3, Table S7). We visually inspected the residuals of the linear
models.

@@

We simulated extinction sequences in the metaweb according to three scenarios to
get expectations on how the food web structure should change according to the
addition and deletion of species (results presented in Fig. 3b) and trophic
species (results in Appendix), i.e. species subdivided in size classes. The
scenario of extinction sequences were related to the trophic level of the
species/trophic species. In the decreasing and increasing scenario, we removed
species/nodes in decreasing, resp. increasing order, of their trophic level,
i.e. respectively beginning by the highest and lowest trophic level. In the
random extinction sequence, we built 50 random extinction sequences. For each
simulation, we removed fish species/nodes that lost all their prey. We then
computed the connectance, the average and the maximum trophic level after each
extinction.

@@

Consumers gain biomass by eating other species and lose biomass by being
consumed and by metabolic loses ($x_i$, eq. \@ref(eq:cons)). The functional
response (eq. \@ref(eq:holling)) of consumer feeding depends on the preference
for its preys ($\omega_{ij}$, set even across consumer preys) and
half-saturation rate ($B_0$, set to 0.5). The hill coefficient ($h$, set to 1.5)
determines the shape of the functional response (Holling's type II or III).
